OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-01-22 paper/paper.Rmd
Draft results & discussion

@@

@@

- TODO: Describe OptiClust and vsearch.

@@

- TODO: brief description of OptiFit algorithm. the method we came up with is the result. OptiClust paper had toy example to walk through the algorithm. 
- TODO: Paragraph at beginning to describe datasets: why picked, how processed.

We used four published datasets isolated from soil
[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse
[@schloss_stabilization_2012], and human [@baxter_microbiota-based_2016]
samples.

- 16S rRNA gene sequence datasets from human gut, mouse gut, marine, and soil environments as well as the greengenes, silva, and RDP public databases were processed with mothur and clustered _de novo_ with mothur's OptiClust algorithm.
- All clustering was performed at a distance threshold of 0.03 and repeated with 100 different random seeds.
- The Matthews Correlation Coefficient was calculated to serve as a measure of OTU quality. OTU quality was evaluated using the Matthews Correlation
Coefficient (MCC) with a sequence dissimilarity threshold of 0.03% as described
previously [@schloss_application_2016; @westcott_novo_2015].

@@

- Public reference databases were clustered _de novo_ using OptiClust, then query datasets were fit to the _de novo_ OTUs using OptiFit.
  - In open-reference mode, OTU quality was similar between fitting the datasets to reference OTUs with OptiFit and clustering the datasets _de novo_ with OptiClust. This held true for all datasets and reference databases.
  - However, in closed-reference mode, OTU quality was slightly worse when fitting to greengenes and silva, and much worse when fitting to RDP as compared to OptiClust. OptiFit was able to map more query sequences to reference OTUs created with the greengenes and silva databases than with RDP.
  - In terms of runtime, closed-reference OptiFit outperformed OptiClust, while OptiClust out-performed open-reference OptiFit.

@@

- TODO: compare optifit performance for split dataset vs public database
- TODO: double-check with Sarah that MCCs are from full OTU dataset or just query sequnce OTUs
- TODO: double-check reference fractions aren't flipped (fraction mapped plot looks weird)

- Datasets were randomly split into a reference fraction and a query fraction. Reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query. Reference sequences were clustered _de novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with OptiFit.
  - OTU quality from fitting split datasets was highly similar to that from _de novo_ clustering the whole dataset.
  - Closed-reference OptiFit with split datasets was faster than OptiClust on whole datasets.
    - OptiClust performed faster than open-reference OptiFit only when the OptiFit reference fraction was 30% or less.
  - Different methods for selecting the sequences to be used as the reference were tested; a simple random sample, weighting sequences by relative abundance, and weighting by similarity to other sequences in the dataset.
    - OTU quality was similar with the simple and abundance-weighted sampling, but slightly worse with similarity-weighted sampling.
    - The fraction of query sequences that are able to be fit to the reference OTUs in closed-reference mode decreases as the reference fraction increases.

### Comparison to vsearch

- TODO: move this into first section with public database.
- vsearch is more permissive than mothur (radius vs diameter with 0.03 dissimilarity threshold).

- To compare to existing software, vsearch was used to cluster OTUs _de novo_ or with reference-based clustering to the greengenes database.
  - For all datasets and clustering methods (_de novo_, open reference, and closed reference), mothur's clustering algorithms produced higher quality OTUs than vsearch.
  - When closed-reference clustering against the greengenes database, vsearch was able to map more query sequences to the reference than mothur's OptiFit algorithm.
  - In terms of runtime, OptiFit genearlly performed faster than vsearch when reference clustering, while vsearch _de novo_ clustering outperformed OptiClust.

@@

@@

### Sequence Data Processing Steps

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
bash. Dependencies include mothur v1.45.0 [@schloss_introducing_2009], vsearch
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
environment manager [@noauthor_anaconda_2016]. A reproducible version of the
workflow, manuscript, and conda environment is available at 
**TODO: UPDATED REPO LINK**.

>>>>

- **TODO: summarize results & conclusion**

@@

**TODO**

@@

- **TODO: broad context; importance of amplicon sequencing**
- **TODO: describe reference vs de novo clustering. trade-offs. machine learning use-case.**
- **TODO: mention OptiClust and vsearch.**

@@

### OptiFit algorithm

OptiFit leverages the method employed by OptiClust of iteratively assigning sequences to OTUs to produce the highest quality OTUs possible, and extends this method for reference-based clustering.

- **TODO: brief description of OptiFit algorithm**
- **TODO: open & closed reference modes**
- **TODO: mcc score; only considers query seqs (when printref=f)**

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. There are two strategies for
generating OTUs with OptiFit: 1) fit sequences to reference OTUs of an
independent database, or 2) split the dataset into a reference and query
fraction, then fit the query sequences to OTUs generated by clustering the
reference sequences _de novo_. For each dataset repeated with 100 random seeds,
we generated OTUs with OptiFit using both strategies, and also clustered _de
novo_ OTUs with OptiClust for comparison. All clustering was performed at a
sequence dissimilarity threshold of 0.03 and OTU quality was evaluated using the
Matthews Correlation Coefficient (MCC) as described previously
[@schloss_application_2016; @westcott_novo_2015]. We calculated the fraction of
query sequences that were fit to existing OTUs in closed-reference mode as an
additional measure of quality for this mode.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database (v13_8_99), Silva non-redundant database (v132), and the Ribosomal
Database Project (RDP; v16). In open-reference mode, OTU quality was similar
between fitting the datasets to reference OTUs with OptiFit and clustering the
datasets _de novo_ with OptiClust. This held true for all datasets and reference
databases. However, in closed-reference mode, OTU quality was slightly worse
when fitting to greengenes and silva, and much worse when fitting to RDP as
compared to OptiClust. No more than half of query sequences were fit to
reference OTUs in closed-reference mode across any dataset/database combination.
OptiFit was able to fit more query sequences to reference OTUs created with the
greengenes and silva databases than with RDP. In terms of runtime,
closed-reference OptiFit outperformed OptiClust, while OptiClust outperformed
open-reference OptiFit.

To compare to existing software, vsearch was used to cluster OTUs _de novo_ or
with reference-based clustering to the greengenes database. For all datasets and
clustering methods (_de novo_, open reference, and closed reference), mothur's
clustering algorithms produced higher quality OTUs than vsearch. When
closed-reference clustering against the greengenes database, vsearch was able to
map more query sequences to the reference than mothur's OptiFit algorithm. In
terms of runtime, OptiFit generally performed faster than vsearch when reference
clustering, while vsearch _de novo_ clustering outperformed OptiClust.

@@

Datasets were randomly split into a reference fraction and a query fraction.
Reference sizes from 10% to 80% of the sequences were created, with the
remaining sequences used for the query. Reference sequences were clustered _de
novo_ with OptiClust, then query sequences were fit to the _de novo_ OTUs with
OptiFit.

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode. OTU quality was remarkably stable across 100 different random seeds. In
terms of runtime, closed-reference OptiFit performed faster than OptiClust on
whole datasets. In open-reference mode, OptiClust performed faster than OptiFit
only when the OptiFit reference fraction was 30% or less. The split dataset
strategy performed just as well as the database strategy in open-reference mode
regardless of database used, and outperformed the database strategy in
closed-reference mode.

We also tested three methods for selecting the sequences to be used as the
reference; a simple random sample, weighting sequences by relative abundance,
and weighting by similarity to other sequences in the dataset. OTU quality was
similar with the simple and abundance-weighted sampling, but slightly worse with
similarity-weighted sampling. In closed-reference mode, The fraction of query
sequences that can be fit to the reference OTUs decreases as the reference
fraction increases.

@@

We developed a new algorithm for fitting sequences to existing OTUs and have
demonstrated its suitability for reference-based clustering. OptiFit makes the
iterative method employed by OptiClust available for tasks where
reference-based clustering is required. We have shown that OTU quality is
similar between OptiClust and OptiFit in open-reference mode, regardless of
strategy employed. Open-reference OptiFit does perform slower than OptiClust
due to the additional _de novo_ clustering step, so users may prefer OptiClust
for tasks that do not require reference OTUs.

When fitting to public databases, OTU quality dropped in closed-reference mode
to different degrees depending on the database and dataset source, and no more
than half of query sequences were able to be fit to OTUs across any
dataset/database combination. This may reflect limitations of reference
databases, which are unlikely to contain sequences from rare and novel microbes.
This drop in quality was most notable with RDP. We recommend users who require
an independent reference database opt for Greengenes or Silva instead. Since
OptiClust performs faster than open-reference OptiFit and creates higher quality
OTUs than closed-reference OptiFit with the database strategy, we recommend
using OptiClust rather than fitting to a database where possible.

The mothur algorithms produced higher quality OTUs than vsearch in
open-reference, closed-reference, or _de novo_ modes. However, vsearch was able
to fit more sequence into OTUs than OptiFit in closed-reference mode. While both
mothur and vsearch use a dissimilarity threshold for determining how to assign
sequences into OTUs, vsearch is more permissive than mothur. Mothur requires
that all pairs of sequences in an OTU are within the dissimilarity threshold
without penalizing the MCC, while vsearch only requires sequences to be similar
to one other sequence in the OTU. In this way, vsearch sacrifices OTU quality in
order to allow more sequences to fit to OTUs. Users who require closed-reference
clustering may prefer to use vsearch if they wish to maximize the fraction of
sequences that can be fit at the cost of OTU quality. However, mothur's
OptiClust or OptiFit are preferred for _de novo_ or open-reference clustering.

When fitting with the split dataset strategy, OTU quality was remarkably similar
when reference sequences were selected by a simple random sample or weighted by
abundance, but quality was slightly worse when sequences were weighted by
similarity. We recommend using a simple random sample since the more
sophisticated reference selection methods do not offer any benefit. The
similarity in OTU quality between OptiClust and OptiFit with this strategy
demonstrates the suitability of using OptiFit to fit sequences to existing OTUs,
such as when using already-trained machine learning models to make predictions
on new data.

- **TODO: big picture concluding paragraph**

@@

### Data Processing Steps

@@

We implemented the analysis workflow in Snakemake [@koster_snakemake_2012] and
relied on Python [@van_rossum_python_2009], R [@r_core_team_r_2020], and GNU
bash. Software used includes mothur v1.45.0 [@schloss_introducing_2009], vsearch
v2.13.3 [@rognes_vsearch_2016], numpy [@harris_array_2020], the Tidyverse
metapackage [@wickham_welcome_2019], R Markdown [@xie_r_2018], and the conda
environment manager [@noauthor_anaconda_2016]. The complete workflow,
manuscript, and conda environment are available at **TODO: UPDATED REPO LINK**.
